---
title: "EDA"
output: html_document
date: "2023-06-26"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(sf)
library(randomForest)
library(vip)
county_health_data_clean <- read_csv("Dropbox/sure2023/data/county_health_data_clean.csv")

county_health_data_clean$MentalHealthProviders <- county_health_data_clean$MentalHealthProviders * 100000

county_health_data_clean %>% 
  filter(CountyRanked == 1) %>%
  head()
```

```{r MHPxPMHD}
county_health_data_clean %>%
  filter(CountyRanked == 1) %>%
  ggplot(aes(x = MentalHealthProviders, y = PoorMentalHealthDays, 
             )) + 
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(trans = "log10")+
  theme_bw() +
  labs(x = "Mental Health Providers Per 100k People", y = "Poor Mental Health Days", 
       title = "Scatterplot of Poor Mental Health Days vs Mental Health Providers",
       subtitle = "County-Level Data",
       color = "Poor Mental Health Days")
```

```{r}
county_health_data_clean %>%
  ggplot(aes(x = MentalHealthProviders, y = PoorMentalHealthDays,
             color = SomeCollege)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_gradient(low = "orange", high = "blue") +
  theme_bw() +
  scale_x_continuous(trans = "log10") +
  labs(x = "Mental Health Providers per 100k People", y = "Poor Mental Health Days", 
       title = "Scatterplot of Mental Health Providers vs. Poor Mental Health Days",
       subtitle = "County-Level Data",
       color = "% College Graduates")
```

```{r}
library(ggplot2)
county_health_data_clean %>%
  ggplot(aes(x = SomeCollege, y = PoorMentalHealthDays,
             color = SomeCollege)) + 
  geom_point(alpha = .5) +
  scale_color_gradient(low = "orange", high = "blue") +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw() +
  labs(x = "% College Graduates", y = "Poor Mental Health Days", 
       title = "Scatterplot of Poor Mental Health Days vs % College Graduates",
       subtitle = "County Level Data",
       color = "% College Graduates")
```

```{r}
# ACA as a factor
county_health_data_clean$ACA <- factor(
  county_health_data_clean$ACA,
  levels = c("Not Implemented", "Implemented")
)

select_states <- c("FL", "AL", "NY", "LA")
county_health_data_clean %>% 
  filter(State %in% select_states, CountyRanked == 1) %>%
  mutate(OrderedState = factor(State, levels = select_states)) %>%
  ggplot(aes(x = MentalHealthProviders, fill = ACA)) + 
  geom_histogram(bins = 30) +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(OrderedState ~ ., ncol = 2) +
  labs(x = "Mental Health Providers per 100k People", 
       y = "Count", 
       title = "Mental Health Providers per 100k People, ACA vs. Non-ACA States",
       subtitle = "County Level Data at Four Select States",
       color = "ACA")
```

```{r}
select_states <- c("FL", "AL", "NY", "LA")
county_health_data_clean %>% 
  # select counties belonging to the states
  filter(CountyRanked == 1, State %in% select_states) %>%
  # fix the order of states
  mutate(OrderedState = factor(State, levels = select_states)) %>%
  # plot
  ggplot(aes(x = MentalHealthProvidersPer100K, y = PoorMentalHealthDays,
             color = ACA)) + 
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() + 
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(trans = "log10") +
  facet_wrap(. ~ OrderedState, ncol = 2) + 
  labs(x = "Number of Mental Health Providers Per 100K Population",
       y = "Number of Poor Mental Health Days",
       title = "Poor Mental Health Days vs. Mental Health Providers, ACA vs. Non-ACA States",
       subtitle = "County Level Data at Four Select States")
```

```{r}
lookup <- c(MedianHouseholdIncome = "Median Household Income", SomeCollege = "Some College")

county_health_data_clean %>%
  select(PoorMentalHealthDays, SomeCollege, MedianHouseholdIncome) %>%
  pivot_longer(cols = -PoorMentalHealthDays,
               names_to = "Name",
               values_to = "Value") %>%
  mutate(Name = recode(Name, !!!lookup)) %>%
  ggplot(aes(x = Value, y = PoorMentalHealthDays)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(. ~ Name, scales = "free_x") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(
    x = "Socioeconomic status",
    y = "Poor Mental Health Days",
    title = "Poor Mental Health Days VS. Socioeconomic status",
    subtitle = "County Health Data"
  )
```

```{r}
county_health_data_clean %>%
  ggplot(aes(x = MedianHouseholdIncome, y = PoorMentalHealthDays,
             color = MedianHouseholdIncome)) +
  geom_point(alpha = .5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_gradient(low = "orange", high = "blue") +
  theme_bw() +
  labs(x = "Median Household Income", y = "Poor Mental Health Days", 
       title = "Median Household Income vs Poor Mental Health Days",
       subtitle = "County Level Data", 
       color = "Median Household Income")
```

```{r}
county_health_data_clean %>% 
  filter(CountyRanked == 1) %>%
  ggplot(aes(x = MentalHealthProvidersPer100K, y = PoorMentalHealthDays,
             color = ACA)) + 
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() + 
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(trans = "log10") +
  facet_grid(. ~ ACA) +
  labs(x = "Number of Mental Health Providers Per 100K Population",
       y = "Number of Poor Mental Health Days",
       title = "Poor Mental Health Days vs. Mental Health Providers Per 100K Population",
       subtitle = "Counties in ACA vs. Non-ACA States")
```

```{r}
lm_socio <- lm(PoorMentalHealthDays ~ MentalHealthProviders +
                 SomeCollege + MedianHouseholdIncome,
               data = county_health_data_clean)
summary(lm_socio)
```


We now fit a random forest model using several predictors in the dataset.
We remove counties that have any missing entry in any of the predictors. 
The following shows the list of predictors used as well as the number of missing entries for each.

```{r, echo = FALSE}
county_health_data_RF <- county_health_data_clean %>% 
  select(PoorMentalHealthDays, LifeExpectancy, LongCommuteDrivingAlone,
         DrivingAlonetoWork, SevereHousingProblems,
         SocialAssociations, ChildreninSingleParentHouseholds, IncomeInequality,
         Unemployment, SomeCollege, HighSchoolCompletion, MentalHealthProviders,
         Uninsured, TeenBirths, SexuallyTransmittedInfections, ExcessiveDrinking,
         AccesstoExerciseOpportunities, FoodEnvironmentIndex, State, PoorPhysicalHealthDays,
         ACA, MedianHouseholdIncome)

colSums(is.na(county_health_data_RF))
```

```{r, echo = FALSE}
# remove missing data rows
county_health_data_RF <- county_health_data_RF %>% drop_na()

# Fit the model
rf_model <- randomForest(PoorMentalHealthDays ~ LongCommuteDrivingAlone +
                         SevereHousingProblems + SocialAssociations +
                         ChildreninSingleParentHouseholds + 
                         IncomeInequality + Unemployment + SomeCollege + 
                         MentalHealthProviders + 
                         Uninsured + TeenBirths + 
                         ExcessiveDrinking + AccesstoExerciseOpportunities + 
                         FoodEnvironmentIndex + 
                         ACA + MedianHouseholdIncome, 
                         data = county_health_data_RF, importance = TRUE, 
                         ntree = 5000)

rf_model
```

```{r}
#variable importance
vip(rf_model) + theme_minimal() + 
  labs(title = "Variable Importance Plot, Random Forest",
       y = "Importance (% Increase in MSE)")
```

The predictor `MentalHealthProviders` has the following variable importance.

```{r}
# % increase in MSE for MentalHealthProviders
importance(rf_model)["MentalHealthProviders", 1]
```

It is not considered more important than some of the variables we considered.
This makes sense because variables such as excessive drinking, unemployment, and socioeconomic status are more direct predictors of mental health problems than the availability of mental health providers.
Nevertheless, the number of mental health providers is still useful in accurately predicting the number of poor mental health days.



